Hydra battles in Coq

### 2020-09-25 doc/hydras.tex
various infix operators for ordinal inequalities

@@

For upgrading constants and fonctions of \texttt{T1}, we have to prove that 
the term they build is in normal form.
For instance, let us represent the ordinals $0$ and $\omega$ as instances of the class \texttt{E0}.

@@

The following lemma establishes the connection between the function
\texttt{succ} and the predicate \texttt{is\_succ}.

@@

Example e'2 : omega o< omega + 6.
Proof. now compute. Qed.

@@

Example e'''2 : omega o< omega * 6.
Proof. now compute. Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma plus_not_monotonous : exists alpha beta gamma : T1,
    alpha o< beta /\ alpha + gamma = beta + gamma.
Proof.
  exists 3, 5, omega;  now  compute.
Qed.

Lemma mult_not_monotonous :  exists alpha beta gamma : T1,
      alpha o< beta /\ alpha * gamma = beta * gamma.
Proof.
  exists 3, 5, omega; now compute.
Qed.
\end{Coqsrc}

@@

  beta : T1
  n : nat
  gamma : T1
  IHbeta : nf beta -> Acc LT beta
  IHgamma : nf gamma -> Acc LT gamma
  H : nf (ocons beta n gamma)
  delta : T1
  Hdelta : delta o< ocons beta n gamma
  ============================
  Acc LT delta
 \end{Coqanswer}

@@

  alpha : T1
  Aalpha : forall y : T1,  y o< alpha -> Acc LT y
  IHalpha : forall y : T1,
       LT y alpha ->
       forall (n : nat) (beta : T1),
       nf (ocons y n beta) -> Acc LT (ocons y n beta)
  ============================
   forall (n : nat) (beta : T1),
   nf (ocons alpha n beta) -> Acc LT (ocons alpha n beta)
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma oplus_strict_mono_LT_l (alpha beta gamma : T1) :
  nf gamma   -> alpha  o< beta ->
  alpha o+ gamma o< beta o+ gamma.

Lemma oplus_strict_mono_LT_r (alpha beta gamma : T1) :
  nf alpha -> beta o< gamma ->
  alpha o+ beta o< alpha o+ gamma.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma S0_decr :
  forall s s', S0  s s' -> ms s' o< ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R1_decr : forall h h',
                  R1 h h' -> m h' o< m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma S1_decr n:
  forall s s', S1 n s s' -> ms s' o<  ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R2_decr n : forall h h', R2 n h h' -> m h'  o< m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma round_decr : forall h h', h -1-> h' -> m h' o< m h.
Proof.
   destruct 1 as [n [H | H]].
   -  now apply R1_decr.
   -  now apply R2_decr with n.
Qed.
\end{Coqsrc}

@@

Theorem every_battle_terminates: Termination.
Proof. 
  red; apply Inclusion.wf_incl with 
         (R2 := fun h h' =>  m h o< m h').
   red; intros;  now apply round_decr.
   apply Inverse_Image.wf_inverse_image, T1_wf.
Qed.
\end{Coqsrc}

@@

\label{lemma:canonS_LT}
\begin{Coqsrc}
Lemma canonS_LT i alpha :
  nf alpha -> alpha <> zero -> canonS alpha i o<  alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma canonS_limit_strong (lambda : T1) : 
     nf lambda ->
     limitb lambda  ->
     forall beta, beta o< lambda ->
                  {i:nat | beta o< canonS lambda i}.

@@

\begin{Coqsrc}
Lemma LT_path_to (alpha beta : T1) :
  beta o< alpha -> {s : list nat | path_to beta s alpha}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma path_to_LT beta s alpha :
  path_to beta s alpha -> nf alpha -> beta o< alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_to_battle alpha beta :
    beta o< alpha ->  iota alpha -+-> iota beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Class BoundedVariant (B:Battle) :=
  {
    mu:T1 ;
    m: Hydra -> T1;
    mu_nf: nf mu;
    Hvar: Hvariant T1_wf B m;
    m_bounded: forall h, m h o< mu
  }.
\end{Coqsrc}

@@

  Hypothesis m_decrease : forall  i h h',
        round_n i h h' -> m h' o< m h.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma m_ge_0 alpha:  nf alpha -> alpha o<= m (iota alpha).
\end{Coqsrc}

@@

\begin{Coqsrc}
    Corollary m_ge_generic : m big_h o<= m small_h.
 \end{Coqsrc}

The (big) rest of the proof is dedicated to prove formally the converse inequality 
\texttt{m small\_h o< m big\_h}.

@@

\begin{Coqsrc}
Lemma m_ge : m big_h o<= m small_h.
  Proof.
    apply m_ge_generic.
   (* ... *)
\end{Coqsrc}

\item From the hypothesis \texttt{m\_bounded}, we have \texttt{m big\_h o< mu}
\item By Lemma \texttt{LT\_to\_battle}, we get a (free) battle from
\texttt{big\_h = iota mu} to \texttt{small\_h = iota (m big\_h)}.

@@

\begin{Coqsrc}
Lemma m_lt : m small_h o< m big_h.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_to_round_plus alpha beta :
    beta o< alpha ->  iota alpha -+-> iota beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Corollary Cor12 (alpha : T1) :  nf alpha ->
         forall beta i n, beta  o< alpha  ->
                (i < n)%nat ->
                 const_pathS i alpha beta ->
                 const_pathS n alpha beta.
Proof.
  transfinite_induction_lt alpha.
  (* (long) proof skipped *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Corollary Cor12' (alpha : T1) :  nf alpha ->
         forall beta i n, (beta o< alpha)%t1 ->
               (i <= n)%nat ->
               const_pathS i alpha beta ->
               const_pathS n alpha beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma Lemma2_6_1 (alpha : T1) :  
  nf alpha -> forall beta,  beta o< alpha  ->
  {n:nat | const_pathS n alpha beta}.
Proof.
  transfinite_induction alpha.
  (* rest of proof skipped *)
\end{Coqsrc}

@@

Lemma constant_to_standard_path 
  (alpha beta : T1) (i : nat):
  nf alpha -> const_pathS i alpha beta -> zero  o< alpha ->
  {l:nat | standard_path (S i) alpha j beta}.
\end{Coqsrc}

@@

  Variables (alpha beta: T1) (n : nat).
  Hypotheses (Halpha: nf alpha) (Hpos : zero o<  beta)
             (p : const_pathS n alpha  beta).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma constant_to_standard_path 
  (alpha beta : T1) (i : nat):
  nf alpha -> const_pathS i alpha beta -> zero  o< alpha ->
  {j:nat | standard_path (S i) alpha j beta}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Corollary  LT_to_standard_path  (alpha beta : T1) :
  beta o< alpha ->
  {n : nat & {j:nat | standard_path (S n) alpha j beta}}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_to_standard_battle :
    forall alpha beta,
      beta o< alpha ->
      exists n i,  battle standard  n (iota alpha) i (iota beta).
\end{Coqsrc}

@@

\begin{Coqsrc}
 Lemma m_ge : m big_h o<= m small_h.
\end{Coqsrc}

>>>>

\begin{todo}
Infix notation for inequalities on type \texttt{T1} have changed to \texttt{t1<} and \texttt{t1<=}.  This document is not updated yet, but the \coq{} scripts are correct.
\end{todo}

@@

\index{Exercises}
\begin{exercise}
In earlier versions of this development, the predicate \texttt{nf} was defined  inductively, with various constructors describing all possible cases.
\begin{enumerate}
\item Please give such a definition, in a dedicated module.
\item Prove the logical equivalence between your definition and ours.
\item Define a variant of the type \texttt{E0} (with your definition of \texttt{nf}).
\item Can you still prove a lemma like \texttt{E0\_eq\_iff} ? With the help of an axiom of the standard library ?
\end{enumerate}
\end{exercise}
For upgrading constants and fonctions of \texttt{T1}, we have to prove that 
the term they build is in normal form.
For instance, let us represent the ordinals $0$ and $\omega$ as instances of the class \texttt{E0}.

@@

The following lemma establishes the connection between the  function
\texttt{succ} and the Boolean predicate \texttt{succb}.

@@

Example e'2 : omega t1< omega + 6.
Proof. now compute. Qed.

@@

Example e'''2 : omega t1< omega * 6.
Proof. now compute. Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma plus_not_monotonous : exists alpha beta gamma : T1,
    alpha t1< beta /\ alpha + gamma = beta + gamma.
Proof.
  exists 3, 5, omega;  now  compute.
Qed.

Lemma mult_not_monotonous :  exists alpha beta gamma : T1,
      alpha t1< beta /\ alpha * gamma = beta * gamma.
Proof.
  exists 3, 5, omega; now compute.
Qed.
\end{Coqsrc}

@@

  beta : T1
  n : nat
  gamma : T1
  IHbeta : nf beta -> Acc LT beta
  IHgamma : nf gamma -> Acc LT gamma
  H : nf (ocons beta n gamma)
  delta : T1
  Hdelta : delta t1< ocons beta n gamma
  ============================
  Acc LT delta
 \end{Coqanswer}

@@

  alpha : T1
  Aalpha : forall y : T1,  y t1< alpha -> Acc LT y
  IHalpha : forall y : T1,
       LT y alpha ->
       forall (n : nat) (beta : T1),
       nf (ocons y n beta) -> Acc LT (ocons y n beta)
  ============================
   forall (n : nat) (beta : T1),
   nf (ocons alpha n beta) -> Acc LT (ocons alpha n beta)
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma oplus_strict_mono_LT_l (alpha beta gamma : T1) :
  nf gamma   -> alpha  t1< beta ->
  alpha o+ gamma  t1< beta o+ gamma.

Lemma oplus_strict_mono_LT_r (alpha beta gamma : T1) :
  nf alpha -> beta t1< gamma ->
  alpha o+ beta t1< alpha o+ gamma.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma S0_decr :
  forall s s', S0  s s' -> ms s' t1< ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R1_decr : forall h h',
                  R1 h h' -> m h' t1< m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma S1_decr n:
  forall s s', S1 n s s' -> ms s' t1<  ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R2_decr n : forall h h', R2 n h h' -> m h'  t1< m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma round_decr : forall h h', h -1-> h' -> m h' t1< m h.
Proof.
   destruct 1 as [n [H | H]].
   -  now apply R1_decr.
   -  now apply R2_decr with n.
Qed.
\end{Coqsrc}

@@

Theorem every_battle_terminates: Termination.
Proof. 
  red; apply Inclusion.wf_incl with 
         (R2 := fun h h' =>  m h t1< m h').
   red; intros;  now apply round_decr.
   apply Inverse_Image.wf_inverse_image, T1_wf.
Qed.
\end{Coqsrc}

@@

\label{lemma:canonS_LT}
\begin{Coqsrc}
Lemma canonS_LT i alpha :
  nf alpha -> alpha <> zero -> canonS alpha i t1<  alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma canonS_limit_strong (lambda : T1) : 
     nf lambda ->
     limitb lambda  ->
     forall beta, beta t1< lambda ->
                  {i:nat | beta t1< canonS lambda i}.

@@

\begin{Coqsrc}
Lemma LT_path_to (alpha beta : T1) :
  beta t1< alpha -> {s : list nat | path_to beta s alpha}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma path_to_LT beta s alpha :
  path_to beta s alpha -> nf alpha -> beta t1< alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_to_battle alpha beta :
    beta t1< alpha ->  iota alpha -+-> iota beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Class BoundedVariant (B:Battle) :=
  {
    mu:T1 ;
    m: Hydra -> T1;
    mu_nf: nf mu;
    Hvar: Hvariant T1_wf B m;
    m_bounded: forall h, m h t1< mu
  }.
\end{Coqsrc}

@@

  Hypothesis m_decrease : forall  i h h',
        round_n i h h' -> m h' t1< m h.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma m_ge_0 alpha:  nf alpha -> alpha t1<= m (iota alpha).
\end{Coqsrc}

@@

\begin{Coqsrc}
    Corollary m_ge_generic : m big_h t1<= m small_h.
 \end{Coqsrc}

The (big) rest of the proof is dedicated to prove formally the converse inequality 
\texttt{m small\_h t1< m big\_h}.

@@

\begin{Coqsrc}
Lemma m_ge : m big_h t1<= m small_h.
  Proof.
    apply m_ge_generic.
   (* ... *)
\end{Coqsrc}

\item From the hypothesis \texttt{m\_bounded}, we have \texttt{m big\_h t1< mu}
\item By Lemma \texttt{LT\_to\_battle}, we get a (free) battle from
\texttt{big\_h = iota mu} to \texttt{small\_h = iota (m big\_h)}.

@@

\begin{Coqsrc}
Lemma m_lt : m small_h t1< m big_h.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_to_round_plus alpha beta :
    beta t1< alpha ->  iota alpha -+-> iota beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Corollary Cor12 (alpha : T1) :  nf alpha ->
         forall beta i n, beta  t1< alpha  ->
                (i < n)%nat ->
                 const_pathS i alpha beta ->
                 const_pathS n alpha beta.
Proof.
  transfinite_induction_lt alpha.
  (* (long) proof skipped *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Corollary Cor12' (alpha : T1) :  nf alpha ->
         forall beta i n, (beta t1< alpha)%t1 ->
               (i <= n)%nat ->
               const_pathS i alpha beta ->
               const_pathS n alpha beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma Lemma2_6_1 (alpha : T1) :  
  nf alpha -> forall beta,  beta t1< alpha  ->
  {n:nat | const_pathS n alpha beta}.
Proof.
  transfinite_induction alpha.
  (* rest of proof skipped *)
\end{Coqsrc}

@@

Lemma constant_to_standard_path 
  (alpha beta : T1) (i : nat):
  nf alpha -> const_pathS i alpha beta -> zero  t1< alpha ->
  {l:nat | standard_path (S i) alpha j beta}.
\end{Coqsrc}

@@

  Variables (alpha beta: T1) (n : nat).
  Hypotheses (Halpha: nf alpha) (Hpos : zero t1<  beta)
             (p : const_pathS n alpha  beta).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma constant_to_standard_path 
  (alpha beta : T1) (i : nat):
  nf alpha -> const_pathS i alpha beta -> zero  t1< alpha ->
  {j:nat | standard_path (S i) alpha j beta}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Corollary  LT_to_standard_path  (alpha beta : T1) :
  beta t1< alpha ->
  {n : nat & {j:nat | standard_path (S n) alpha j beta}}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_to_standard_battle :
    forall alpha beta,
      beta t1< alpha ->
      exists n i,  battle standard  n (iota alpha) i (iota beta).
\end{Coqsrc}

@@

\begin{Coqsrc}
 Lemma m_ge : m big_h t1<= m small_h.
\end{Coqsrc}
